\documentclass[10pt, a4paper]{article}
\usepackage{preamble}

\title{Discrete Mathematics Problem Sheet Solutions}
\author{Luke Phillips}
\date{October 2024}

\begin{document}

\maketitle

\newpage

\section{Introduction}

\begin{problem}[Q10]\phantom{}
    \begin{enumerate}[label = (\alph*)]
        \item In how many ways can $6$ people be seated around dinner table if all rotations of a given arrangement are regarded as the same?
        \item In how many ways can $6$ different spherical beads be arranged on a circular ring?
    \end{enumerate}
    \begin{enumerate}[label = (\alph*)]
        \item 
        \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
            If we assign a number to each seat and a letter to each person then person $A$ has $6$ choices, person $B$ has $5$ then person $C$ has $4$ choices and so on until there is only one seat left. This looks like the following
            \[
            6 \times 5 \times 4 \times 3 \times 2 \times 1 = 6!,
            \]
            hence there are $6!$ ways of seating $6$ people around the table, however, for each way there is a rotation which is similar therefore we would divide by $6$ to get $\frac{6!}{6} = 5!$ ways excluding rotations.
        \end{proof}
        \item
        \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
            This is similar to the previous problem as there are $6$ different beads and they can be arranged in different ways. We have to exclude rotations by the definition of just looking at the circular ring. Therefore we get $5!$ ways of arranging the beads.
        \end{proof}
    \end{enumerate}
\end{problem}

\newpage

\begin{problem}[Q11]
    \begin{enumerate}[label = (\alph*)]
        \item How many $5$ letter words can be made from $\{\text{X, Y, Z}\}$ that include at least one $X$?
        \item How many of the words specified in (a) also contain Y?
        \item How many of the words specified in (b) also contain Z?
    \end{enumerate}

    \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
        
    \end{proof}
\end{problem}

\newpage

\begin{problem}[Q12]\phantom{}
    \begin{enumerate}[label = (\alph*)]
        \item In how many ways can $8$ identical pieces be placed on the squares of an $8$-by-$8$ chess board so that no row or column contains more than one piece?
        \item The same question as in (a) but with $n$ pieces on an $n$-by-$n$ board.
    \end{enumerate}
    \begin{enumerate}[label = (\alph*)]
        \item 
        \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
            To simplify this problem we will think of all of the different positions along the row for each piece on each row. Firstly, the first piece has $8$ options to be placed on, but since this eliminates one of the columns, the second piece (the piece on the row above the first) will only have $7$ options. This pattern continues, where the third piece only has $6$ options, the fourth $5$ options, and so on until we reach the eighth piece with only one option. Therefore using the multiplication principle we can see that the total number of ways is
            \[
            8 \times 7 \times 6 \times 5 \times 4 \times 3 \times 2 \times 1 = 8!.
            \]
        \end{proof}
        \item 
        \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
            Using similar logic to that of the previous question we can see that, for an $n\times n$ grid the choices go as follows,
            \begin{enumerate}[label = {}]
                \item $n$ choices for the $1^{\text{st}}$ row.
                \item $n - 1$ choices for the $2^{\text{nd}}$ row.
                \item $\vdots$
                \item $2$ choices for the $(n - 1)^{\text{th}}$ row.
                \item $1$ choice for the $n^{\text{th}}$ row.
            \end{enumerate}
            We can clearly see that the number of ways will therefore be $n!$.
        \end{proof}
    \end{enumerate}
\end{problem}

\newpage

\begin{problem}[Q14]
    A class consists of $n$ girls and $m$ boys. In how many ways can they stand in a single line so that all the girls are together?

    \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
        
    \end{proof}
\end{problem}

\newpage

\begin{problem}[Q15]
    How many rearrangements of the letters FORMULA contain FOR as three successive letters in that order? How about FOR together but not necessarily in that order?
    \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
        When we are considering rearrangements of the letters of FORMULA which contain FOR as three successive letters we can see that there are $7$ letters in the word FORMULA and $3$ of them have to be taken by FOR hence there are $5$ blocks (or boxes) which can contain either a letter or FOR. The remaining $4$ letters can be placed in the blocks in $4!$ different ways. This means that there are $5$ different ways to place the FOR in the blocks and $4!$ different ways to place the rest of the letters in the blocks, hence the total number of rearrangements is
        \[
        5 \cdot 4! = 120.
        \]
        If we consider FOR still being together but not in that order we get
        \[
        5 \cdot 4! \cdot 3! = 720
        \]
        rearrangements as there are $3!$ ways of rearranging FOR when it is in its block.
    \end{proof}
\end{problem}

\newpage

\begin{problem}[Q16]
    $6$ people sit on a row of $6$ chairs. They stand up and then sit down again (possibly on different chairs). In how many ways can the re-seating be done so that the two end chairs are each not occupied by the same person as sat on them originally.
    \begin{proof}[Solution]\renewcommand{\qedsymbol}{}
        
    \end{proof}
\end{problem}

\newpage

\section{Revision lecture}

\begin{problem}
    $G(V, E)$ connected simple planar,
    every vertex degree $3$,
    every face with either $4, 6$ edges,
    exactly $2$ faces with $6$ edges.
    How many faces of $4$ edges are there?
    Draw an example.

    \begin{solution}
        Euler's theorem:
        \[
        V - E + F = k + 1.
        \]
        Since complete $2k$ connected components
        \[
        V - E + F = 2k + 1.
        \]
        
        Handshaking lemma:
        \[
        \sum_{G}d(v) = 2|E|
        \]
        \[
        \sum_{f}b(f) = 2|E|.
        \]

        Two faces with six edges
        \[
        2|E| = 2 \cdot 6 + 4(F \cdot 2)
        \]

        since every vertex is of degree $3$:
        \[
        3V = 2E
        \]
        from the handshaking lemma.

        \[
        3V = 2E = 2 \cdot 6 + 4(F \cdot 2)
        \]
        \[
        2V - 2E + 2F = 4 \iff -V + 2F = 4
        \]
        \begin{align*}
            -3V &= 12 - 6F \\
            3V &= 12 + 4F - 8 \\
            &\iff \\
            0 &= 24 - 2F - 8 \implies 2F = 16 \implies F = 8.
        \end{align*}

        There are $6$ faces with $4$ edges
        (as there are two with six edges so $F - 2 = 6$).

        Diamond inside of a diamond with each edge on the smaller diamond connecting the the respective one on the larger one.
        Then there is one face of $6$ on the inside diamond and one on the exterior.
        
    \end{solution}
\end{problem}

\begin{problem}
    Let $K_n$ be the complete graph on $n \geq 3$ edges.
    \begin{enumerate}[label = (\alph*)]
        \item For which values of $n$ is $K_n$ Hamiltonian?
        
        \item For which values of $n$ is $K_n$ Eulerian?
        
        \item For which values of $n$ is $K_n$ Planar?

        \item Show that $K_{2n}$ is a union of $n$ edge disjoint walks in $K_{2m}$.
    \end{enumerate}

    \begin{solution}
        \begin{enumerate}[label = (\alph*)]
            \item $n \geq 3$.
            Take $C_n = (v_1, v_2, \dotsc, v_n, v_1)$ is a Hamiltonian cycle.

            \item $K_n$ is Eulerian if and only if every vertex has even degree if and only if $n$ is odd.

            \item 
            By some theorem,
            $G$ is planar if and only if it contains no subgraph which is a subdivision of $K_5$ or $K_{3, 3}$.
            If $n \geq 5$ then $K_5$ is a subset of $K_n$ and so $K_n$ is not planar.

            \item
            The union of $G = (V, E)$ and $G' = (V', E')$ is the graph $\tilde{G} = (\tilde{V}, \tilde{E})$ where $\tilde{V} = V \cup V'$ and $\tilde{E} = E \cup E'$.
            Consider $K_{2n} \subset K_{2n + 1}$.
            By Euler's theorem there is an Eulerian cycle $W$ in $K_{2n + 1}$
            $W = (v_{2n + 2}, w_1, w_2, \dotsc, v_{2n + 1}, w_1, \dotsc, v_{2n + 1})$ so
            \[
            \#\{i \in \N\,:\,w_i = v_{n + i}\} = \frac{2m - 2}{2} + 2 = m + 1.
            \]
            \[
            \{i \in \N\,:\,w_i = v_{2n + i}\} = \{0, i_1, i_2, \dotsc, i_n\}
            \]
            for $j \leq m$ let $W ^ {(j)} = (w_{i_{j - 1}}, w_{i_{j - 1} + 2}, \dotsc, w_{i_{j - 1}})$.
            $W ^ {(j)} \subseteq K_{2n}$ because $W$ spans
            (uses)
            all the edges we have $\bigcup_{j = 1}W ^ {(j)} = K_{2n}$.
            $W ^ {(j)} \cap W ^ {(j + 1)} = \emptyset$ by edge disjointness of Eulerian circuit.
        \end{enumerate}
    \end{solution}
\end{problem}














\end{document}