\documentclass[10pt, a4paper]{article}
\usepackage{preamble}

\title{Dynamics I}
\author{Luke Phillips}
\date{January 2025}

\begin{document}

\maketitle

\newpage

\tableofcontents

\newpage

\section{Particle on a line}

\subsection{The Basics}

\begin{definition}[Position]
    The position $x(t)$ where $t$ is time.
\end{definition}

\begin{definition}[Velocity]
    The velocity $\dot{x}(t)$ where $t$ is time.
\end{definition}

\textit{Note that $\dot{x}$ means $\frac{dx}{dt}$.}

\begin{definition}[Acceleration]
    The acceleration $a(t) = \dot{v}(t) = \ddot{x}(t)$.
\end{definition}

\begin{definition}[Speed]
    The speed is $|v(t)|$,
    the magnitude of the velocity.
\end{definition}

\begin{definition}[Mass]
    A particle has mass $m > 0$.
\end{definition}

\begin{definition}[Momentum]
    A particle's momentum,
    $p$ is given by
    \[
    p = mv = m\dot{x}.
    \]
\end{definition}

\begin{definition}[Equation of motion]
    In an applied force $F$,
    then the equation of motion is $\dot{p} = F$.
\end{definition}

\begin{remark}\phantom{}
    \begin{enumerate}[label = (\roman*)]
        \item Could have $m$ depending on time
        (e.g. rocket).

        Unless otherwise stated,
        we take $m$ to be a \textbf{constant}.

        Then the EoM\footnote{Equation of motion} is $m\dot{v} = m\ddot{x} = F$.

        \item In general,
        $F = F(t, x, \dot{x})$ and $m\ddot{x} = F(t, x, \dot{x})$ is a second order differential equation which cannot be solved
        (explicitly).
    \end{enumerate}
\end{remark}

We focus on special cases which can be solved.

The simplest example is $F$ is constant.
For example,
vertical motion under gravity close to the surface of a planet:
then the magnitude of $F$ is $|F| = mg$
(the weight)
where the constant $g$ depends on the planet.

\begin{example}
    An object is shot up vertically from the ground with some initial speed $u$.
    What maximum height $H$ does it reach?
    \begin{solution}
        Could use $x$ upwards or $x$ downwards,
        both work,
        choose a convenient one.
        Here having $x$ upwards seems more natural.
        Then $m\ddot{x} = -mg$ with $x = 0$ at the ground.
        Then
        \begin{align*}
            \ddot{x} = -g &\xrightarrow{\int\,dt} \dot{x} = -gt + u \\
            \intertext{where $u$ is the constant of integration.} \\
            &\xrightarrow{\int\,dt} x(t) = -\frac{1}{2}gt ^ 2 + ut + 0.
        \end{align*}
        Max $\dot{x} = 0 \implies t = u / g$.
        So $H = x\left(\frac{u}{g}\right) = -\frac{1}{2}g\left(\frac{u}{g}\right) ^ 2 + u\left(\frac{u}{g}\right) = \frac{u ^ 2}{2g}$.
    \end{solution}
\end{example}

\subsection{Case \texorpdfstring{$F = F(t)$}{} and \texorpdfstring{$F = F(\dot{x})$}{}}
The same method works,
do $\int\,dt$ twice.

\begin{example}
    A particle of mass $m = 1$ is at rest at $x = 0$.
    A force $F = e ^ {-t}$ is switched on at time $t = 0$.
    Find $x(t)$ for $t > 0$.

    \textit{Note:
    expect $\dot{x} > 0$ and $x > 0$ for $t > 0$
    (since $F > 0$).}

    \begin{solution}
        \[
        \ddot{x} = e ^ {-t} \xrightarrow{\int\,dt} \dot{x} = -e ^ {-t} + 1 \xrightarrow{\int\,dt} x = e ^ {-t} + t - 1.
        \]
    \end{solution}
\end{example}

\begin{remark}
    Here $m\ddot{x} = F(\dot{x})$ second-order non linear differential equation.
\end{remark}

\begin{example}
    A particle of mass $m$ is slowed by a frictional force of magnitude $be ^ {av}$
    ($a$ and $b$ are constant).
    Initial speed $u$.
    What time $T$ does the particle travel for before coming to rest?
    
    \begin{solution}
        The particle either moves to the right or the left,
        the force moves in the opposite direction.

        Choosing the first situation.
        So
        \begin{align*}
            m\frac{dv}{dt} = -be ^ {av} &\implies \int e ^ {av}\,dv = -\frac{b}{m}\int\,dt \\
            \intertext{could use a constant of integration,
            fixed by $v(0) = u$.} \\
            \int e ^ {av}\,dv = -\frac{b}{m}\int\,dt &\implies \int_{u}^{0} e ^ {av}\,dv = -\frac{b}{m}\int_{0}^{T}\,dt \\
            &\implies \left[-\frac{1}{a}e ^ {-av}\right]_{u}^{0} = -\frac{bT}{m} \\
            &\implies T = \frac{m}{ab}\left(1 - e ^ {-au}\right).
        \end{align*}
    \end{solution}
\end{example}

\begin{example}
    A mass $M$ falls downwards under gravity,
    a resistive force $bw$,
    where $w$ is its speed.
    Initially $w(0) = 0$,
    find $w(t)$.
    What is the "least upper bound" of this $w$,
    the terminal speed.

    \begin{solution}
        We could use the $x$ axis pointing up or down.
        Use $x$ as downwards since the motion is downwards.

        Then
        \[
        m\dot{v} = mg - bv
        \]
        where $v$ is the velocity,
        here the velocity $v = w$.
        So
        \[
        m\dot{w} = mg - bw \implies \frac{dw}{dt} = g - \frac{b}{m}w \implies \int\frac{1}{g - \frac{b}{m}w}\,dw = \int\,dt
        \]
        which results in $w(t) = \frac{mg}{b}\left(1 - e ^ {-\frac{bt}{m}}\right)$,
        the terminal speed is
        \[
        \liminfty[t]\frac{mg}{b}\left(1 - e ^ {-\frac{bt}{m}}\right) = \frac{mg}{b}.
        \]
    \end{solution}
\end{example}

\subsection{Case \texorpdfstring{$F = F(x)$}{}}
Here $m\ddot{x} = F(x)$.
If $F(x) = bx + c$
($b$ and $c$ are constant),
then $m\ddot{x} - bx = c$ linear,
the solution is the complementary function $+$ the particular integral.
This doesn't work for $F(x)$ non-linear.
For that think of $v = v(x)$.
Then
\[
m\frac{dv}{dt} = m\frac{d}{dt}v(x(t)) = m\frac{dv}{dx}\cdot\frac{dx}{dt} = m\frac{dv}{dx}v = F(x).
\]
This is a first order separable ODE,
solve to get $v(x)$.
Integrating gives
\[
\frac{1}{2}mv ^ 2 = \int F(x)\,dx,
\]
and hence you get $v(x)$.
To get $x(t)$,
use $\frac{dx}{dt} = v(x)$ again first order separable.

\begin{example}
    A mass with $m = 2$ moves in a force with $F = 3x ^ 2$,
    initial conditions $x = 1$ and $v = 1$ at $t = 0$.
    What time $T$ does it take to reach $x = 9$?

    \begin{solution}
        \begin{align*}
            2v\frac{dv}{dx} = 3x ^ 2 &\implies \int 2v\,dv = \int 3x ^ 2\,dx \\
            &\implies v ^ 2 = x ^ 3 + c &\text{($c = 0$ from initial conditions)} \\
            &\implies v ^ 2 = x ^ 3 \\
            &\implies v = x ^ {\frac{3}{2}} &\text{(positive since $x = v = 1$ at $t = 0$)}
        \end{align*}
        Now
        \begin{align*}
            \frac{dx}{dt} = x ^ {\frac{3}{2}} &\implies \int_{1}^{9} x ^ {-\frac{3}{2}}\,dx = \int_{0}^{T}\,dt \\
            &\implies T = \left[-2x ^ {-\frac{1}{2}}\right]_{2}^{9} = -2\left(\frac{1}{3} - 1\right) = \frac{4}{3}.
        \end{align*}
    \end{solution}
\end{example}

\begin{remark}
    For problems with $F = F(v)$ we can use $F(v) = m\frac{dv}{dt}$ as we saw before.
    Sometimes it may be more convenient to use $F(v) = mv\frac{dv}{dx}$,
    in situations where we want $v(x)$ rather than $v(t)$.
\end{remark}

\subsection{Two more general examples}

\begin{example}['Mixed' example]
    A unit mass moves along a line,
    acted on by a force $F(v, t) = -v + e ^ {-t}$ where $v$ is its velocity.
    Given that $v(0) = 0$,
    find $v(t)$.
    \begin{solution}
        The equation of motion is
        \[
        \dot{v} = -v + e ^ {-t},
        \]
        which is a linear ODE for $v(t)$.
        Solving it via an integrating factor $e ^ t$ gives
        \[
        v(t) = te ^ {-t}.
        \]
    \end{solution}
\end{example}

\begin{example}[Rocket: non-constant mass]
    A rocket of mass $m(t)$ burns fuel at a constant rate $\dot{m} = -c$,
    where $c > 0$.
    This produces a constant thrust $F = k|\dot{m}| = kc$.
    The rocket starts at rest at $x = 0$,
    with initial mass $m_0$.
    Find its subsequent position $x(t)$.
    \begin{solution}
        The constant force gives $mv = kct$.
        Now $m(t) = m_0 - ct$,
        so
        \[
        v(t) = \frac{kct}{m_0 - ct} = -k + \frac{km_0}{m_0 - ct}
        \]
        for $0 \leq t < \frac{m_0}{c}$.
        Integrating this gives
        \[
        x(t) = -kt + \frac{km_0}{c}\log\left(\frac{m_0}{m_0 - ct}\right)
        \]
        for $0 \leq t < \frac{m_0}{c}$.
        Note that for small $t$ we have
        \[
        v(t) \approx \frac{kct}{m_0};
        \]
        whereas when $t \rightarrow \frac{m_0}{c}$ we get $v \rightarrow \infty$ and $x \rightarrow \infty$.
    \end{solution}
\end{example}

\subsection{Energy}

\begin{definition}
    Consider the case $F = F(x)$,
    and write $V(x) = -\int F(x)\,dx$.
    Then
    \[
    \frac{d}{dx}\left(\frac{1}{2}mv ^ 2\right) = mv\frac{dv}{dx} = ma = F = -\frac{dV}{dx},
    \]
    and so
    \[
    E = \frac{1}{2}mv ^ 2 + V(x)
    \]
    is constant in time,
    since the chain rule gives
    \[
    \frac{dE}{dt} = \frac{dE}{dx}\dot{x} = 0.
    \]
    This quantity $E$ is the energy of the particle,
    we say that it is conserved.
    The term $K = \frac{1}{2}mv ^ 2$ is the kinetic energy,
    $V(x)$ is the potential energy;
    neither of these is conserved in general.
\end{definition}

\begin{remark}
    \begin{enumerate}[label = (\roman*)]
        \item A force of the form $F(x)$ is said to be conservative,
        since it conserves energy.

        \item The potential $V$,
        and hence the energy $E$,
        are only defined up to an arbitrary constant.

        \item The simplest example is where $F$ is constant,
        so $V$ is linear in $x$.

        \item Using conservation of energy can give us a good picture of the particle's motion,
        even when we cannot get $x(t)$ explicitly.

        \item A point $x$ where $F(x) = 0$,
        or equivalently $V'(x) = 0$,
        is called an equilibrium position.
    \end{enumerate}
\end{remark}

\begin{example}
    A unit-mass particle moves along the $x$-axis,
    subject to a force $F = -2x ^ 3$.
    Derive an expression for its energy $E$,
    such that $E = 0$ if the particle is stationary at $x = 0$.
    Then compute $E$ if its speed is $v = 1$ at $x = 1$.
    \begin{solution}
        Writing $V(x) = -\int F(x)\,dx$.
        \[
        V(x) = -\int -2x ^ 3\,dx = \frac{1}{2}x ^ 4 + C.
        \]
        Now we have
        \begin{align*}
            E &= \frac{1}{2}mv ^ 2 + V(x) \\
            &= \frac{1}{2}v ^ 2 + \frac{1}{2}x ^ 4 + C \\
            \intertext{But $E = 0, v = 0, x = 0$ so $0 = 0 + 0 + C$ hence $C = 0$} \\
            &= \frac{1}{2}\left(v ^ 2 + x ^ 4\right).
        \end{align*}
        Finally,
        we can see that $E$ with $v = 1$ and $x = 1$ is
        \[
        E = \frac{1}{2}\left(v ^ 2 + x ^ 4\right) = \frac{1}{2}(1 + 1) = 1.
        \]
    \end{solution}
\end{example}

\subsection{Motion in a potential}
Consider a particle moving along the $x$-axis,
in a potential $V(x)$.
Except in simple cases,
we won't know exactly what the trajectory $x(t)$ is.
But we can say something about what it looks like,

\begin{example}
    A double-well potential.
    A particle of mass $3$ moves in a potential
    \[
    V(x) = \frac{x ^ 4}{4} - \frac{x ^ 3}{3} - x ^ 2.
    \]
    At what points can the particle remain at rest
    (be in equilibrium)?
    If the particle starts at $x = 1$ with speed $u$,
    how large does $u$ have to be for it to reach $x = -1$?
    \begin{solution}
        We can see that the equilibrium points are at $V'(x) = 0$.
        Now
        \[
        0 = V'(x) = x(x - 2)(x + 1)
        \]
        so there are three equilibria,
        at $x = -1, 0, 2$.
        The potential $V(x)$ is a 'double well',
        with local maximum at $x = 0$,
        and local minima at $x = -1$ and $x = 2$.
        To get from $x = 1$ to $x = -1$,
        the particle needs enough initial kinetic energy to get over the 'hill' at $x = 0$.
        In other words,
        we need
        \[
        \frac{3}{2}u ^ 2 + V(1) > V(0),
        \]
        using $V(1) = -\frac{13}{12}$ and $V(0) = 0$ then gives $u > \sqrt{\frac{13}{18}}$.
    \end{solution}
\end{example}

\subsection{Forcing and resonance}
Say our car is travelling on a corrugated road surface.
This is modelled by adding a forcing term on the right-hand side,
such as $C\sin(pt)$.
Let us first look at a simple example with no damping,
namely
\[
m\ddot{x} + kx = m\sin(pt).
\]
The general solution is,
\[
x(t) = A\cos(\omega t) + B\sin(\omega t) + \frac{\sin(pt)}{\omega ^ 2 - p ^ 2},
\]
provided $|p| \neq \omega$.
We have a mixture of two oscillations with angular frequencies $\omega$ and $p$.
Note $x(t)$ is bounded.
If $p = \omega$,
the solution is
\[
x(t) = A\cos(\omega t) + B\sin(\omega t) - \frac{t\cos(\omega t)}{2\omega},
\]
This is resonance:
the applied force is synchronised with the natural oscillation of the mass-spring system.
The steady-state response is a periodic particular integral $x_{PI}$,
of the form $x(t) = A\cos(pt) + B\sin(pt)$.
Defining $\phi$ by $\tan(\phi) = \frac{2bp}{\omega ^ 2 - p ^ 2}$,
$x_{PI}$ can be written in the form $x_{PI}(t) = C\sin(pt - \phi)$.
The phase difference $\phi$ between forcing and response is due to a lag between when the car feels an upward force from the uneven road,
and when it actually moves up.
\begin{example}
    The behaviour $x(t)$ of a forced damped system is described by the equation $\ddot{x} + 4\dot{x} + x - \sin(t) = 0$.
    Find the steady-state response $x_S(t)$ of this system,
    and compute the phase difference between it and the external force.
    \begin{solution}
        The steady-state response is the periodic particular integral of the differential equation,
        and this is
        \begin{align*}
            x_S(t) = -\frac{1}{4}.
        \end{align*}
        Hence from
        \[
        \tan(\phi) = \frac{2bp}{\omega ^ 2 - p ^ 2}
        \]
        we get
        \[
        \tan(\phi) = \frac{2b}{1 ^ 2 - 1 ^ 2} \implies \phi = \frac{\pi}{2}.
        \]
        The phase difference between this and $\sin(t)$ is $\frac{\pi}{2}$
    \end{solution}
\end{example}

\subsection{Small Oscillations}
Motion near stable equilibria.
A conservative force $F(x) = -V'(x)$ has an equilibrium at $x = x_0$ if $F(x_0) = 0$.
A particle can sit at $x = x_0$,
at rest.
The equilibrium is stable if $x_0$ gives a local minimum of $V$,
so $V''(x_0) > 0$.
If,
on the other hand,
$V''(x_0) \leq 0$,
then the equilibrium is unstable.

\textbf{Simplest case}.
The simplest case is $V(x) = \frac{1}{2}kx ^ 2$,
with $k > 0$
(the simple harmonic oscillator).
There is a stable equilibrium at $x = 0$.
Then $m\ddot{x} = -kx$,
and the system oscillates with angular frequency $\omega = \sqrt{\frac{V''(0)}{m}} = \sqrt{\frac{k}{m}}$.
The period if $\frac{2\pi}{\omega} = 2\pi\sqrt{\frac{m}{k}}$.

\textbf{General potential}.
Suppose that $V(x)$ has a local minimum at $x_0$.
Write $x(t) = x_0 + \epsilon(t)$.
Thus we get approximate simple harmonic motion,
with angular frequency approximately $\omega = \sqrt{\frac{V''(x_0)}{m}}$ and period $\frac{2\pi}{\omega}$.
Note that $\epsilon$ remains small,
so this makes sense.
In effect,
we are approximating the graph of $V(x)$ by a parabola near $x_0$.

\subsection{Momentum and Collisions}
\textbf{Conservation of momentum}.
If there are no external forces,
and $N$ particles with momenta $p_1, p_2, \dotsc, p_N$,
then the total momentum
\[
p = p_1 + p_2 + \dotsi + p_{N}
\]
is conserved,
and does not change when the particles collide.

A collision which conserves energy is said to be elastic,
one which doesn't is inelastic.

\begin{example}
     A particle of mass $2m$ and speed $v$ collides head-on with a particle of mass m at rest.
     The two then continue to move in the same direction,
     with the heavier particle moving at speed $\frac{v}{2}$.
     What is the speed $u$ of the lighter particle,
     and what fraction of the initial energy is lost in the collision?

     \begin{solution}
         By conservation of momentum:
         \[
         2mv = 2m\frac{v}{2} + mu \iff u = v.
         \]
         Now,
         the initial energy was $E_0 = \frac{1}{2}(2m)v ^ 2 = mv ^ 2$,
         the final energy is $E_1 = \frac{1}{2}(2m)\left(\frac{v}{2}\right) ^ 2 + \frac{1}{2}(m)v ^ 2 = \frac{3}{4}mv ^ 2$.
         Hence the fraction of the initial energy lost in the collision is
         \[
         \frac{E_0 - E_1}{E_0} = \frac{mv ^ 2 - \frac{3}{4}mv ^ 2}{mv ^ 2} = 1 - \frac{3}{4} = \frac{1}{4}.
         \]
     \end{solution}
\end{example}

\section{Dynamics in Space}

\subsection{Velocity and relative motion}
A particle is an object with mass $m$,
located at a point in $3$D space.
Its position is described by a position vector $\mbf{r}$
(relative to some origin $\mbf{O}$).
We can use Cartesian coordinates $(x, y, z)$ by letting $\mbf{e}_1, \mbf{e}_2, \mbf{e}_3$ be the standard unit vectors,
and writing $\mbf{r} = x\mbf{e}_1 + y\mbf{e}_2 + z\mbf{e}_3$.
This system of axes is an example of a frame of reference.
For motion in a plane,
use $\mbf{r} = x\mbf{e}_1 + y\mbf{e}_2$.

The trajectory of a particle is given by $\mbf{r}(t) = x(t)\mbf{e}_1 + y(t)\mbf{e}_2 + z(t)\mbf{e}_3$.
Its velocity is $\mbf{v} = \dot{\mbf{r}}$,
where as usual the dot denotes the derivative with respect to time $t$.
If $\mbf{r} = x\mbf{e}_1 + y\mbf{e}_2 + z\mbf{e}_3$,
then $\mbf{v} = \dot{x}\mbf{e}_1 + \dot{y}\mbf{e}_2 + \dot{z}\mbf{e}_3$.
The speed of the particle $v = |\mbf{v}|$ is a real number with $v \geq 0$.
The acceleration is $\mbf{a} = \dot{\mbf{v}} = \ddot{\mbf{r}}$.

\begin{example}
    If $\mbf{r} = t\mbf{e}_1 + \mbf{e}_2 + t ^ 2\mbf{e}_3$ what is $\mbf{v}, \mbf{a}$ and $v$?

    \begin{solution}
        \begin{align*}
            \mbf{v} &= \dot{\mbf{r}} = \mbf{e}_1 + 2t\mbf{e}_3 \\
            \mbf{a} &= \dot{\mbf{v}} = 2\mbf{e}_3 \\
            v &= \sqrt{1 ^ 2 + (2t) ^ 2} = \sqrt{1 + 4t ^ 2}.
        \end{align*}
    \end{solution}
\end{example}

The momentum of a particle with mass $m$ and velocity $\mbf{v}$ is the vector $\mbf{p} = m\mbf{v}$.
If an external force $\mbf{F}$ acts on the particle,
then its equation of motion is $\dot{\mbf{p}} = \mbf{F}$.
If $m$ is constant,
this is equivalent to $m\ddot{\mbf{r}} = \mbf{F}$.

If particles $A$ and $B$ are located at $\mbf{r}_A$ and $\mbf{r}_B$ respectively,
then the position of $B$ relative to $A$ is $\mbf{r}_B - \mbf{r}_A$.
If they are moving with velocities $\mbf{v}_A$ and $\mbf{v}_B$ respectively,
then the velocity of $B$ relative to $A$ is $\mbf{v}_B - \mbf{v}_A$.

\begin{example}
    Two particles have positions $\mbf{r}_A = 2t\mbf{e}_1 + \mbf{e}_2 + t ^ 2\mbf{e}_3$ and $\mbf{r}_B = -t\mbf{e}_1$ respectively,
    what is their relative position and velocity?
    
    \begin{solution}
        The relative position is
        \[
        \mbf{r}_B - \mbf{r}_A = -t\mbf{e}_1 - 2t\mbf{e}_1 - \mbf{e}_2 - t ^ 2\mbf{e}_3 = -3t\mbf{e}_1 - \mbf{e}_2 - t ^ 2\mbf{e}_3.
        \]
        The relative velocity is
        \begin{align*}
            \mbf{v}_B - \mbf{v}_A &= \dot{\mbf{r}_B} - \dot{\mbf{r}_A} \\
            &= (-\mbf{e}_1) - (2\mbf{e}_1 + 2t\mbf{e}_3) \\
            &= -\mbf{e}_1 - 2\mbf{e}_1 - 2t\mbf{e}_3 \\
            &= -3\mbf{e}_1 - 2t\mbf{e}_3.
        \end{align*}
    \end{solution}
\end{example}

\begin{example}
    Bob is going north
    (direction $\mbf{e}_2$)
    with speed $10$,
    and feels a headwind of speed $25$.
    What is the wind velocity relative to the ground?
    If Bob now goes east
    (direction $\mbf{e}_1$)
    at speed $15$,
    what wind does he feel?

    \begin{solution}
        \[
        \mbf{v}_w - \mbf{v}_B = -25\mbf{e}_2 \iff \mbf{v}_w = -15\mbf{e}_2,
        \]
        so the wind velocity relative to the ground is $-15\mbf{e}_2$.

        Now we have that $\mbf{v}_B = 15\mbf{e}_1$,
        so
        \[
        \mbf{v}_w - \mbf{v}_B = -15\mbf{e}_2 - 15\mbf{e}_1
        \]
        hence he feels a speed of
        \[
        |\mbf{v}_w - \mbf{v}_B| = |-15\mbf{e}_2 - 15\mbf{e}_1| = \sqrt{15 ^ 2 + 15 ^ 2} = 15\sqrt{2}
        \]
        in the direction $-\frac{(\mbf{e}_1 + \mbf{e}_2)}{\sqrt{2}}$.
    \end{solution}
\end{example}

\subsection{Sliding down an inclined plane}
A mass $m$ is free to slide down a plane which is inclined at an angle $\alpha$ to the horizontal.
Gravity acts downwards,
there is a frictional force
(directed up the slope)
with magnitude $\mu \mbf{N}$,
where $\mu$ is a positive constant,
and $\mbf{N}$ is the normal force exerted by the inclined plane on the particle.
(The direction of $\mbf{N}$ is perpendicular to the plane).
Starting from rest,
how far does the particle travel in time $t$?
How large does $\alpha$ have to be for the mass to move at all?

\begin{solution}
    Take $\mbf{e}_1$ to point down the slope.
    Take $\mbf{e}_2$ to be perpendicular to $\mbf{e}_1$.

    Motion is in the $\mbf{e}_1$ direction.
    The weight of the particle is downwards,
    with magnitude $mg$.
    Hence its component in the $\mbf{e}_2$-direction is $-mg\cos{\alpha}$,
    and so $\mbf{N} = mg(\cos{\alpha})\mbf{e}_2$.
    Thus,
    the total force in the $\mbf{e}_1$ direction is
    \[
    F = -\mu mg\cos{\alpha} + mg\sin{\alpha}
    \]
    which is constant;
    so
    \[
    x(t) = \frac{1}{2}g(\sin{\alpha} - \mu\cos{\alpha})t ^ 2,
    \]
    provided $\tan{\alpha} > \mu$,
    else the particle does not move,
    and $x(t) = 0$ for all $t$.
\end{solution}

\subsection{Motion in an electromagnetic field}

A particle with mass $m$ and electric charge $q$,
placed in an electric field $\mbf{E}$ and a magnetic field $\mbf{B}$,
feels the Lorentz force
\[
\mbf{F} = q(\mbf{E} + \mbf{v} \times \mbf{B}).
\]

\begin{example}
    Take $\mbf{E} = \mbf{0}$ and $\mbf{B} = B\mbf{e}_3$ with $B$ constant,
    and with initial conditions $\mbf{r}(0) = \mbf{0}$ and $\dot{\mbf{r}}(0) = u\mbf{e}_1$.
    The equation of motion is
    \begin{align*}
        m\ddot{\mbf{r}} &= \mbf{F} \\
        &= q(\mbf{E} + \mbf{v} \times \mbf{B}) \\
        &= q(\mbf{v} \times \mbf{B}) \\
        &= Bq\left(\begin{pmatrix}
            \dot{x} \\ \dot{y} \\ \dot{z}
        \end{pmatrix} \times \begin{pmatrix}
            0 \\ 0 \\ 1
        \end{pmatrix}\right) \\
        &= Bq\begin{pmatrix}
            \dot{y} \\ \dot{x} \\ 0
        \end{pmatrix}
    \end{align*}
    so we have
    \[
    m\begin{pmatrix}
        \ddot{x} \\
        \ddot{y} \\
        \ddot{z}
    \end{pmatrix} = \begin{pmatrix}
        Bq\dot{y} \\
        Bq\dot{x} \\
        0
    \end{pmatrix}
    \]
    which is equivalent to
    \[
    m\ddot{x} = Bq\dot{y} \implies m\dot{x} = Bqy + C
    \]
    \[
    m\ddot{y} = Bq\dot{x} \implies m\dot{y} = -Bqx + D
    \]
    and
    $\ddot{z} = 0 \implies z = 0$.

    Applying the initial conditions $x(0) = 0, \dot{x}(0) = u, y(0) = 0, \dot{y}(0) = 0$
    \[
    m\dot{x}(0) = Bqy(0) + C \iff C = mu
    \]
    and
    \[
    m\dot{y}(0) = -Bq(0) + D \iff D = 0.
    \]
    Hence we get
    \[
    m\dot{x} = Bqy + mu
    \]
    \[
    m\dot{y} = -Bqx.
    \]
    So we have two simultaneous equations,
    solving for $x(t)$
    \begin{align*}
        m\dot{x} &= Bqy + mu \\
        &\implies \\
        m\ddot{x} &= Bq\dot{y} \\
        &= Bq\left(-\frac{Bq}{m}x\right) \\
        &= -\frac{(Bq) ^ 2}{m}x \\
        &\iff \\
        \ddot{x} + \left(\frac{Bq}{m}\right) ^ 2x &= 0 \\
        &\implies \\
        x(t) &= C\sin\left(\frac{Bq}{m}t\right) + D\cos\left(\frac{Bq}{m}t\right)
    \end{align*}
    applying the initial conditions we get
    \[
    x(t) = \frac{Bq}{m}u\sin\left(\frac{Bq}{m}t\right)
    \]
    then applying the equation for $x(t)$,
    integrating and applying initial conditions we get
    \[
    y(t) = \frac{Bq}{m}u\left(\cos\left(\frac{Bq}{m}t\right) - 1\right).
    \]
\end{example}

\subsection{Conservative forces and potential energy}
The kinetic energy of a particle of mass $m$ and speed $v$ is $\frac{1}{2}mv ^ 2$.
If the force $\mbf{F}$ depends only on position $\mbf{r}$,
and if there exists a real function $V(\mbf{r})$ such that
\[
\mbf{F} = \left(-\pd[V]{x}, -\pd[V]{y}, -\pd[V]{z}\right) = -\pd[V]{x}\mbf{e}_1 -\pd[V]{y}\mbf{e}_2 - \pd[V]{z}\mbf{e}_3,
\]
then the force is conservative,
and $V$ is called the potential energy.
The total energy is
\[
E(\mbf{r}, \mbf{v}) = \frac{1}{2}mv ^ 2 + V(\mbf{r}).
\]

\begin{proposition}
    In a conservative force field,
    the total energy is conserved.

    \begin{proof}
        Write $\mbf{v} = (v_x, v_y, v_z)$.
        Then
        \[
        \frac{dV}{dt} = \pd[V]{x}v_x + \pd[V]{y}v_y + \pd[V]{z}v_z = -\mbf{F} \cdot \mbf{v},
        \]
        by the chain rule.
        Secondly,
        \[
        \frac{d}{dt}\left(\frac{1}{2}mv ^ 2\right) = \frac{d}{dt}\left(\frac{1}{2}m(v_x ^ 2 + v_y ^ 2 + v_z ^ 2)\right) = m(v_x\dot{v}_x + v_y\dot{v}_y + v_z\dot{v}_z) = m\mbf{a} \cdot \mbf{v}.
        \]
        And so
        \[
        \frac{dE}{dt} = (m\mbf{a} - \mbf{F}) \cdot \mbf{v} = 0.
        \]
    \end{proof}
\end{proposition}

\begin{remark}
    The force $\mbf{F} = F_x\mbf{e}_1 + F_y\mbf{e}_2 + F_z\mbf{e}_3$ is conservative if and only if
    \[
    \pd[F_z]{y} = \pd[F_y]{z},\qquad\pd[F_x]{z} = \pd[F_z]{x},\qquad\pd[F_y]{x} = \pd[F_x]{y}.
    \]
\end{remark}

\begin{example}
    Take $\mbf{F} = (z - y)\mbf{e}_1 + (\alpha x - z)\mbf{e}_2 + (\beta x - y)\mbf{e}_3$,
    where $\alpha$ and $\beta$ are constants.
    For which value(s) of these constants,
    if any,
    is $\mbf{F}$ conservative?
    For these values,
    find $V$.

    \begin{solution}
        We can start by identifying that
        \[
        \pd[V]{x} = z - y \implies V(x, y, z) = (y - z)x + g(y, z)
        \]
        this is because $\mbf{F} = \left(-\pd[V]{x}, -\pd[V]{y}, -\pd[V]{z}\right)$ so $-\pd[V]{x} = (z - y)$.

        Now we can apply this $V$ since we know that
        \[
        \pd[V]{y} = z - \alpha x = (0 - 1)x + h(z) \iff h(z) - x
        \]
        so $h(z) = z$ and $\alpha = -1$.

        Similarly
        \[
        \pd[V]{z} = y - \beta x = -x + f(y) \iff f(y) - x
        \]
        so $f(y) = y$ and $\beta = 1$.

        Since $\pd{y}g(y, z) = h(z) = z$ and $\pd{z}g(y, z) = f(y) = y$ we have naturally that $g(y, z) = yz$.

        Using these we can see that
        \[
        V(x, y, z) = (y - z)x + yz = yx - xz + yz + C
        \]
        where $C$ is an arbitrary constant.
    \end{solution}
\end{example}

\subsection{The simple pendulum}
A point mass $m$ is attached to one end of a rigid light rod of length $L$.
The rod is pivoted at the other end,
and is free to swing in a vertical plane.
Let $\theta(t)$ denote the angle between the pendulum and the vertical.
The gravitational potential energy is $V = mgz$,
where $z$ is the height of the mass above its hanging-straight-down position.
So $V$ is effectively a function of $\theta$ given by
\[
V(\theta) = mgL(1 - \cos{\theta}).
\]
If the coordinate in the horizontal direction is $x$,
then $x = L\sin{\theta}$,
and so
\[
v ^ 2 = \dot{x} ^ 2 + \dot{z} ^ 2 = L ^ 2\dot{\theta} ^ 2.
\]
Here $\dot{\theta}$ is the angular velocity;
note $v = L|\dot{\theta}|$.
The total energy is conserved is
\[
E = \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 + mgL(1 - \cos{\theta}).
\]
The equation of motion obtained from $\frac{dE}{dt} = 0$ is $\ddot{\theta} = -\left(\frac{g}{L}\right)\sin{\theta}$.
For small oscillations about the stable equilibrium,
the equation becomes $\ddot{\theta} = -\frac{g\theta}{L}$,
so we get simple harmonic motion with angular frequency $\omega = \sqrt{\frac{g}{L}}$.
This is just an approximation for small $\theta$.

\begin{example}
    Release the pendulum from rest in a horizontal position.
    What is the speed $u$ of the mass when it passes the vertical position?

    \begin{solution}
        By conservation of energy
        \[
        E = \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 + mgL(1 - \cos{\theta}).
        \]
        So initial $E$ equals final $E$
        \[
        V\left(\frac{\pi}{2}\right) - \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 = \frac{1}{2}m(L\dot{\theta}) ^ 2 + V(0) \iff V\left(\frac{\pi}{2}\right) = \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 + 0 + V(0)
        \]
        \[
        V\left(\frac{\pi}{2}\right) = mgL = \frac{1}{2}m(L\dot{\theta}) ^ 2 + V(0) = \frac{1}{2}m(L\dot{\theta}) ^ 2
        \]
        \[
        mgL = \frac{1}{2}m(L\dot{\theta}) ^ 2 \iff \sqrt{2gL} = L|\dot{\theta}| = v
        \]
        hence $v = \sqrt{2gL}$.
    \end{solution}
\end{example}

\begin{example}
    If you start at $\theta(0) = 0$,
    with $\dot{\theta}(0) = \omega$,
    how large does $\omega$ have to be for the pendulum to 'loop the loop'?

    \begin{solution}
        We need $E = \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 + V(0) > V(\pi)$ since the energy before the movement has to be larger than the energy at the top of the swing.
        \[
        \frac{1}{2}mL ^ 2\dot{\theta} ^ 2 > V(\pi) = 2mgL \iff \dot{\theta} > 2\sqrt{\frac{g}{L}}
        \]
        but $\dot{\theta}(0) = \omega$ so $\omega > 2\sqrt{\frac{g}{L}}$.
    \end{solution}
\end{example}

\subsection{Projectiles}
Setup.
You fire a projectile at an angle $\theta$ to the horizontal,
with initial speed $u$.
The subsequent motion is in a plane.
Take axes $z$ and $\mbf{e}_3$ upwards,
and $x$ and $\mbf{e}_1$ horizontal,
with no motion in the $y$-direction.
The initial velocity is
\[
\mbf{v}(0) = u\mbf{e}_1\cos{\theta} + u\mbf{e}_3\sin{\theta}.
\]
The forces are $-mg\mbf{e}_3$ for gravity,
and $-\lambda\mbf{v}$ for air resistance.
Take $\mbf{r}(0) = \mbf{0}$.
If there is no resistance,
then we get
\[
\mbf{r}(t) = ut\mbf{e}_1\cos{\theta} + \left(ut\sin{\theta} - \frac{1}{2}gt ^ 2\right)\mbf{e}_3.
\]

\begin{example}
    Suppose that the ground surface is horizontal,
    and that there is no resistance.
    Let $T$ be the time taken for the projectile to land.
    In other words,
    $z(T) = 0$.
    Find the maximum range of the projectile.

    \begin{solution}
        \[
        \begin{pmatrix}
            x(t) \\ 0 \\ 0
        \end{pmatrix} = \mbf{r}(t) = \begin{pmatrix}
            ut\cos{\theta} \\
            0 \\
            ut\sin{\theta} - \frac{1}{2}gt ^ 2
        \end{pmatrix}
        \]
        so $ut\sin{\theta} - \frac{1}{2}gt ^ 2 = 0 \iff t = \frac{2u}{g}\sin{\theta}$.

        Hence $T = \frac{2u}{g}\sin{\theta}$.
        To maximise the range we have
        \[
        x(t) = ut\cos{\theta} = \frac{2u ^ 2}{g}\sin{\theta}\cos{\theta} = \frac{u ^ 2}{g}\sin(2\theta)
        \]
        so $\dot{x} = \frac{2u ^ 2}{g}\cos(2\theta)$ which maximised would be $\dot{x} = 0 = \frac{2u ^ 2}{g}\cos(2\theta) \implies \theta = \frac{\pi}{4}$.

        So for a given $u$ we can maximise the range by $\theta = \frac{\pi}{4}$.
    \end{solution}
\end{example}


\subsection{Momentum and collisions}

\begin{definition}
    Momentum.
    A particle with mass $m$ and velocity $\mbf{v}$ has momentum $\mbf{p} = m\mbf{v}$.
    If a force $\mbf{F}$ is acting,
    then the equation of motion is $\dot{\mbf{p}} = \mbf{F}$.
    In general,
    momentum is not conserved.

    If
    (say)
    the $x$-component of the force is zero,
    then $p_x$ is conserved
    (constant in time).
\end{definition}

\textbf{Collisions}.
If there are no external forces,
then the total momentum is conserved.
If the collision is elastic,
then kinetic energy is conserved too.

\begin{example}
    A particle of uni mass is travelling with velocity $\mbf{v} = 3\mbf{e}_1 + 3\mbf{e}_2$ when it collides with a stationary particle of mass $2$.
    If the unit-mass particle comes away from the collision with velocity $\mbf{u} = -\mbf{e}_1 + \mbf{e}_2$,
    calculate the velocity $\mbf{w}$ of the other particle after the collision,
    and calculate the fractional loss in kinetic energy.

    \begin{solution}
        Total momentum before is $\mbf{p} = 3\mbf{e}_1 + 3\mbf{e}_2$ and total momentum after is $(-\mbf{e}_1 + \mbf{e}_2) + 2\mbf{w}$.
        Conservation says
        $\begin{pmatrix}
            3 \\ 3
        \end{pmatrix} = \begin{pmatrix}
            -1 \\ 1
        \end{pmatrix} + 2\mbf{w} \iff \mbf{w} = \begin{pmatrix}
            2 \\ 1
        \end{pmatrix}$.

        The total kinetic energy before is $K = \frac{1}{2}\sqrt{3 ^ 2 + 3 ^ 2} ^ 2 = 9$ and after $K' = \frac{1}{2}\sqrt{(-1) ^ 2 + 1 ^ 2} ^ 2 + \frac{1}{2}\cdot 2\sqrt{2 ^ 2 + 1 ^ 2} ^ 2 = 6$ so the fractional kinetic energy loss is
        \[
        \frac{K - K'}{K} = \frac{1}{3}.
        \]
    \end{solution}
\end{example}

\newpage

\section{Central Forces and Orbits}

\subsection{Polar basis}
Using polar coordinates $(r, \theta)$,
$x = r\cos{\theta}$ and $y = r\sin{\theta}$ for our basis.
$\mbf{r} = r\mbf{e}_r$,
$\mbf{e}_r$ is the radial unit vector pointing away from the origin,
the tangential unit vector $\mbf{e}_{\theta}$ is the vector perpendicular to $\mbf{e}_r$.

We have
\[
\mbf{e}_r = \mbf{e}_1\cos{\theta} + \mbf{e}_2\sin{\theta}
\]
and
\[
\mbf{e}_{\theta} = -\mbf{e}_1\sin{\theta} + \mbf{e}_2\cos{\theta}
\]
giving these time derivatives
\begin{align*}
    \dot{\mbf{e}}_r &= \frac{d\theta}{dt}\frac{d}{d\theta}\mbf{e}_r = \dot{\theta}(-\mbf{e}_1\sin(\theta) + \mbf{e}_2\cos{\theta}) = \dot{\theta}\mbf{e}_{\theta} \\
    \dot{\mbf{e}}_{\theta} &= \frac{d\theta}{dt}\frac{d}{d\theta}\mbf{e}_{\theta} = \dot{\theta}(-\mbf{e}_1\cos(\theta) - \mbf{e}_2\sin{\theta}) = -\dot{\theta}\mbf{e}_r
\end{align*}
Since $\mbf{r} = r\mbf{e}_r$,
we can get an expression for velocity
(and acceleration)
\[
\mbf{v} = \dot{r}\mbf{e}_r + r\dot{\mbf{e}}_r = \dot{r}\mbf{e}_r + r\dot{\theta}\mbf{e}_{\theta}.
\]

\subsection{Angular momentum}
A particle with position $\mbf{r}$ and momentum $\mbf{p}$ has angular momentum $\mbf{L} = \mbf{r} \times \mbf{p}$ about $\mbf{r} = \mbf{0}$.

If a force $\mbf{F}$ acts at the point with position vector $\mbf{r}$,
then its torque or moment about $\mbf{r} = \mbf{0}$ is $\mbf{T} = \mbf{r} \times \mbf{F}$.
If the mass is constant,
so that $\dot{\mbf{r}} \times \mbf{p} = \mbf{0}$,
the equation of motion stays that the rate of change of angular momentum equals the applied torque:
\[
\frac{d\mbf{L}}{dt} = \mbf{r} \times \dot{\mbf{p}} = \mbf{r} \times \mbf{F} = \mbf{T}.
\]

\begin{proposition}
    Suppose that the force on a particle at position $\mbf{r}$ is always proportional to $\mbf{r}$.
    Then its angular momentum about $\mbf{0}$ is conserved.

    \begin{proof}
        \[
        \frac{d\mbf{L}}{dt} = \frac{d}{dt}(m\mbf{r} \times \mbf{v}) = \mbf{r} \times \mbf{F} = \mbf{0},
        \]
        since $\mbf{F}$ and $\mbf{r}$ are proportional.
        In other words,
        such a force cannot exert torque,
        which is what changes angular momentum.
    \end{proof}
\end{proposition}

\begin{example}
    A particle with constant mass $m$ has trajectory $\mbf{r}(t) = t\mbf{e}_1 + \mbf{e}_2 + t ^ 2\mbf{e}_3$.
    Calculate its angular momentum $\mbf{L}$ about $\mbf{0}$,
    and compute the torque which is affecting its motion.

    \begin{solution}
        We have
        \[
        \mbf{r}(t) = \begin{pmatrix}
            t \\ 1 \\ t ^ 2
        \end{pmatrix}
        \]
        using $\mbf{L} = \mbf{r} \times \mbf{p}$,
        and $\dot{\mbf{p}} = m\mbf{v} = m\dot{\mbf{v}}$ we have
        \begin{align*}
            \mbf{L} &= \mbf{r} \times \mbf{p} \\
            &= \begin{pmatrix}
                t \\ 1 \\ t ^ 2
            \end{pmatrix} \times m\begin{pmatrix}
                1 \\ 0 \\ 2t
            \end{pmatrix} \\
            &= m\begin{pmatrix}
                2t \\ -t ^ 2 \\ -1
            \end{pmatrix}.
        \end{align*}
        Now for our torque we get $\mbf{T} = \mbf{r} \times \mbf{F}$,
        \begin{align*}
            \mbf{T} &= \mbf{r} \times \mbf{F} \\
            &= \mbf{r} \times m\ddot{\mbf{r}} \\
            &= \mbf{r} \times \dot{\mbf{p}} \\
            &= \frac{d\mbf{L}}{dt} \\
            &= m\begin{pmatrix}
                2 \\ -2t \\ 0
            \end{pmatrix}.
        \end{align*}
    \end{solution}
\end{example}

\subsection{Central forces}
\begin{definition}
    A central force based at $\mbf{r} = \mbf{0}$ has the form
    \[
    \mbf{F} = f(r)\mbf{e}_r = r ^ {-1}f(r)\mbf{r}.
    \]
    Note that $|\mbf{F}| = |f|$.
\end{definition}

The force can act towards or away from the origin $\mbf{O}$:
it is attractive if $f(r) < 0$ and repulsive if $f(r) > 0$.
The angular momentum of a particle moving in a central force is constant.

Since $\mbf{L} \cdot \mbf{r}(t) = 0$,
the motion of the particle lies in the plane through $\mbf{O}$ perpendicular to the constant vector $\mbf{L}$,
and we take this to be the $r\theta$-plane.

Our basis is $\{\mbf{e}_r, \mbf{e}_{\theta}, \mbf{e}_3\}$,
where $\mbf{e}_3$ is constant.

The crucial point is that,
because angular momentum is conserved,
the plane does not change with time.

\begin{example}
    Compute the acceleration $g$ due to gravity at the surface of the Earth.
    The mass of the Earth is $6 \times 10 ^ {24}$kg,
    and its radius is $6400$km.

    \begin{solution}
        
    \end{solution}
\end{example}

\subsection{Motion in a central force}
Motion in polar coordinates.
Take $\mbf{L}$ to be in the $z$-direction,
so that $\mbf{L} = L\mbf{e}_3$.
Let $r$ and $\theta$ be polar coordinates in the $xy$-plane.
Recall that
\[
\mbf{r} = r\mbf{e}_r,\quad\dot{\mbf{r}} = \dot{r}\mbf{e}_r + r\dot{\theta}\mbf{r}_{\theta},\quad\ddot{\mbf{r}} = (\ddot{r} - r\dot{\theta} ^ 2)\mbf{e}_r + (2\dot{r}\dot{\theta} + r\ddot{\theta})\mbf{e}_{\theta}.
\]
Note that the angular momentum is
\[
\mbf{L} = mr\mbf{e}_r \times (\dot{r}\mbf{e}_r + r\dot{\theta}\mbf{e}_{\theta}) = mr ^ 2\dot{\theta}\mbf{e}_3,
\]
using $\mbf{e}_r \times \dot{\theta} = \mbf{e}_3$,
and so  $L = mr ^ 2\dot{\theta}$.
The equation of motion $m\ddot{\mbf{r}} = \mbf{F}$ becomes
\[
m(\ddot{r} - r\dot{\theta} ^ 2) = f(r),\quad 2\dot{r}\dot{\theta} + r\ddot{\theta} = 0.
\]
If we multiply the second of these by $r$,
we see it simply says that $r ^ 2\dot{\theta}$ is constant:
it is equivalent to conservation of angular momentum,
and we can replace it by $mr ^ 2\dot{\theta} = L$.
Then use this to eliminate $\dot{\theta}$ from the first
(radial)
equation,
to leave
\[
m\ddot{r} - \frac{L ^ 2}{mr ^ 3} = f(r).
\]
Note that the kinetic is
\[
K = \frac{1}{2}m|\dot{\mbf{r}}| ^ 2 = \frac{1}{2}m(\dot{r} ^ 2 + r ^ 2\dot{\theta} ^ 2) = \frac{1}{2}m\left(\dot{r} ^ 2 + \frac{L ^ 2}{m ^ 2r ^ 2}\right).
\]
Now any central force is conservative,
and the potential energy is
\[
V(r) = -\int f(r)\,dr,
\]
where the constant of integration is usually chosen so that $V(r) \rightarrow 0$ as $r \rightarrow \infty$.
For example,
the gravitational potential energy is $V(r) = -GMm / r$.
That any central is conservative can be proved directly;
but the crucial justification is that the total energy $E = K + V$ is constant.
Indeed,
\[
\frac{dK}{dt} = \dot{r}\left(m\ddot{r} - \frac{L ^ 2}{mr ^ 3}\right),
\]
and $dV / dt = -f(r)\dot{r}$,
so $dE / dt = 0$ is equivalent to the following equation:
\[
m\ddot{r} - \frac{L ^ 2}{mr ^ 3} = f(r).
\]

\subsection{Bound orbits}

A unit-mass particle moves under the influence of the attractive central force $f(r) = -\frac{1}{r ^ 2}$.
It starts at $r = 1$,
with speed $v_0$ and direction perpendicular to $\mbf{r}$.
Show that the particle cannot escape to infinity
(it is in a bound orbit)
if $v_0 ^ 2 < 2$.
In this case,
find the smallest and largest values of the radius $r$.
What value of $v_0$ gives a circular orbit?

\begin{solution}
    Here $V(r) = -\int f(r)\,dr = -\frac{1}{r}$,
    and so the energy is
    \[
    E = \frac{v ^ 2}{2} - \frac{1}{r} = \frac{v_0 ^ 2}{2} - 1,
    \]
    using the initial conditions.
    If the particle reaches infinity we would have $E = \frac{1}{2}v ^ 2 + 0 \geq 0$,
    so it is bound if $E < 0$,
    which gives $v_0 ^ 2 < 2$.

    For the second part,
    we need the angular momentum
    \[
    L = r ^ 2\dot{\theta} = r(r\dot{\theta}) = v_0.
    \]
    Expressing $v$ in polar coordinates and eliminating $\dot{\theta}$ then gives
    \[
    E = \frac{1}{2}\left(\dot{r} ^ 2 + \frac{v_0 ^ 2}{r ^ 2}\right) - \frac{1}{r} = \frac{v_0 ^ 2}{2} - 1.
    \]
    This equation can be written as $\dot{r} ^ 2 + r ^ {-2}Q(r) = 0$,
    where $Q(r) = (2 - v_0 ^ 2)r ^ 2 - 2r + v_0 ^ 2$.
    The maximum and minimum values of $r$ occur when $\dot{r} = 0$,
    giving a quadratic equation $Q(r) = 0$ for $r$.
    The roots are $r = r_1 = 1$ and $r = r_2 = \frac{v_0 ^ 2}{2 - v_0 ^ 2}$.
    Furthermore,
    $r$ is restricted to a range for which $Q(r) \leq 0$.
    Sketching $Q(r)$ reveals three possibilities:
    if $v_0 ^ 2 < 2$ then $r$ lies between $r_1$ and $r_2$,
    a bound orbit;
    if $v_0 = 1$ then $r = 1$ for all times,
    a circular orbit;
    if $v_0 ^ 2 > 2$ then $r_2 < 0$ and $r > r_1$,
    an unbound trajectory.
    In fact,
    $r_2 \rightarrow \infty$ as $v_0 ^ 2 \rightarrow 2$ from below:
    the escape speed is $v_0 = \sqrt{2}$.
\end{solution}

\begin{example}[Escape speed]
    This is a variant of the previous example.
    A projectile is launched from the surface of the Earth
    (radius $R$ and mass $M$),
    with initial speed $v_0$.
    How large does $v_0$ have to be in order that the projectile escapes the Earth's gravitational pull?

    \begin{solution}
        We need
        \[
        E = \frac{1}{2}mv_0 ^ 2 - \frac{GMm}{R} \geq 0 - 0 = 0,
        \]
        and so $v_0 \geq \sqrt{\frac{2GM}{R}}$,
        consistent with the previous example.
    \end{solution}
\end{example}

\begin{example}
    A unit-mass particle is subject to an attractive central force $f(r) = -\frac{1}{r ^ 3}$.
    What is its escape speed $v_0$ if it starts at $r = R$?

    \begin{solution}
        Here we have
        \[
        V(r) = -\int f(r)\,dr = -\int-\frac{1}{r ^ 3}\,dr = -\frac{1}{2r ^ 2}
        \]
        and we have
        \[
        E = \frac{1}{2}mv_0 ^ 2 + V(r) = \frac{v_0 ^ 2}{2} - \frac{1}{2R ^ 2}
        \]
        but $E \geq 0$ hence
        \[
        0 \leq E = \frac{v_0 ^ 2}{2} - \frac{1}{2}\cdot\frac{1}{R ^ 2} \iff v_0 \geq \frac{1}{R}
        \]
        hence the escape speed is $v_0 = \frac{1}{R}$.
    \end{solution}
\end{example}

\subsection{Finding the orbit}
We want to find the orbit of an object in a central force,
in the form $r = r(\theta)$.
It is convenient to use the variable $u = \frac{1}{r}$ rather than $r$,
so we want an equation for $u(\theta)$.
Note that the chain rules gives
\[
\frac{d}{dt} = \frac{Lu ^ 2}{m}\frac{d}{d\theta}.
\]
This enables us to write $\dot{r}$ and $\ddot{r}$ as
\[
\dot{r} = \frac{dr}{dt} = \frac{Lu ^ 2}{m}\frac{d}{d\theta}\left(\frac{1}{u}\right) = -\frac{L}{m}u',\quad \ddot{r} = \frac{d}{dt}\dot{r} = -\frac{L ^ 2u ^ 2}{m ^ 2}u'',
\]
where $u'$ denotes $\frac{du}{d\theta}$ and similarly for $u''$.
Then the equation of motion becomes
\[
u'' + u = -mL ^ {-2}u ^ {-2}f(u ^ {-1}).
\]
This is the orbit equation,
a second-order differential equation for $u(\theta)$.

\begin{example}
    Let us take $f(r) = -\frac{1}{r ^ 2}$,
    namely gravity,
    and $m = 1$.

    \begin{solution}
        We have the standard orbit equation:
        \[
        u'' + u = -mL ^ {-2}u ^ {-2}f(u ^ {-1})
        \]
        which taking $u = \frac{1}{r}$ gets us
        \[
        u'' + u = -\frac{m}{L ^ 2u ^ 2}\cdot\left(-\frac{1}{r ^ 2}\right) = -\frac{r ^ 2}{L ^ 2}\cdot\left(-\frac{1}{r ^ 2}\right) = \frac{1}{L ^ 2}
        \]
        so we get
        \[
        \frac{d ^ 2u}{d\theta ^ 2} + u = \frac{1}{L ^ 2},
        \]
        then using the fact that
        $m ^ 2 + 1 \iff m ^ 2 = -1$ we get the complementary function
        \[
        u(\theta) = A\sin(\theta) + B\cos(\theta)
        \]
        which we can simplify to being only in terms of cosine
        \[
        u(\theta) = C\cos(\theta + \theta_0).
        \]
        Finding the particular integral through
        \[
        u(\theta) = D = \frac{1}{L ^ 2}\quad u'(\theta) = 0\quad u''(\theta) = 0
        \]
        we get
        \[
        u(\theta) = C\cos(\theta + \theta_0) + \frac{1}{L ^ 2}.
        \]
        We can then rotate this in the polar coordinates,
        we may take $\theta_0 = 0$.
        This gets us
        \[
        r(\theta) = \frac{L ^ 2}{1 + \varepsilon\cos(\theta)},
        \]
        where $\varepsilon \geq 0$ is arbitrary.
        If $\varepsilon < 1$,
        then $r$ is bounded
        (since the denominator is never zero),
        and the curve is an ellipse.
        If $\varepsilon > 1$,
        $r$ is unbounded,
        and the curve is a hyperbola.
        $\varepsilon = 1$ is a parabola,
        also unbounded,
        $\varepsilon = 0$ is a circle.
        The parameter $\varepsilon$ is called the eccentricity of the orbit.
        Then the orbit equation becomes
        \[
        \frac{d ^ 2u}{d\theta ^ 2} + u = \frac{1}{L ^ 2}.
        \]
    \end{solution}
\end{example}

\subsection{Circular orbits and stability}

For a circular orbit.
we need an attractive force with $f(r) < 0$.
Let us write $g(u) = -f(1 / u)$,
so $g(u)$ is positive.
The orbit equation becomes
\[
u'' + u = \frac{mg(u)}{L ^ 2u ^ 2}.
\]
A circular orbit $u = b$
(constant)
is possible,
with the angular momentum satisfying $mg(b) = L ^ 2b ^ 3$.
But is this orbit stable?

We will simply test for stability by making a perturbation which keeps the angular momentum fixed.
The perturbation is to replace $u = b$ by $u = b + \varepsilon(\theta)$,
where $\varepsilon(\theta)$ is small;
the question is whether it stays small.
Substituting into the orbit equation gives,
using the Taylor expansion and after some algebra
\[
\varepsilon'' = \left[\frac{bg'(b)}{g(b)} - 3\right]\varepsilon,
\]
where we ignore terms of order $\varepsilon ^ 2$.
For stability we need the coefficient on the right-hand-side to be negative,
in other words
\[
bg'(b) < 3g(b).
\]
Gravity has $g(u) = ku ^ 2$ which satisfies this.
The case $f(r) = -\frac{2}{r ^ 2} - \frac{1}{r ^ 3}$ also admits circular orbits.
But $f(r) = -\frac{1}{r ^ 3}$ does not.

\begin{example}
    Most communication satellites occupy geostationary orbits,
    namely circular orbits above the equator which stay at the same point relative to the rotating Earth.
    Show that only one height $r_0$ is possible for such an orbit.

    \begin{solution}
        We know $\frac{L}{m} = r_0 ^ 2\dot{\theta}$ by the definition of $L$
        (the definition of angular momentum),
        and also $L ^ 2 / m ^ 2 = GMr_0$ for a circular orbit.
        A geostationary orbit has a particular value of $\dot{\theta}$,
        one full rotation per day.
        So we get an equation
        \[
        \frac{L}{m} = r_0 ^ 2\dot{\theta} \implies \frac{L ^ 2}{m ^ 2} = r_0 ^ 4\dot{\theta} ^ 2  \iff GMr_0 = r_0 ^ 4\dot{\theta} ^ 2 \iff \frac{GM}{\dot{\theta} ^ 2} = r_0 ^ 3
        \]
        this determines $r_0$ uniquely.
        Its value is $r_0 = 42000$km,
        corresponding to a height of $35786$km above the Earth.
    \end{solution}
\end{example}





















\end{document}